---
title: "analy_main"
author: "Niamh MacSweeney"
date: "2023-07-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

##Introduction 

This script follows analy_LCGA and uses the LCGA outputs to address the primary aim (and H1 and H2) of our pre-registered study: how exposure to threat in childhood relates to the rate at which female youth progress through pubertal maturation (i.e., pubertal tempo), and how this relates to later internalising difficulties.

Required inputs: Dataframe with internalising, trauma, and puberty class variables, as well as the fixed and random effect variables. We will load in our required dataframe from the analy_LCGA script. 


```{r set up}


#load libraries
library(tidyverse)
library(hrbrthemes) #for plotting
library(stats)
library(lme4)
library(lmerTest)
library(ggpubr) #plots
library(rstatix) #for avova
library(lavaan)
library(lavaanPlot)
library(tidySEM)
library(semPlot)
library(semTable)
library(ggcorrplot)
library(bestNormalize) #for transforming data 
library(gtsummary) #for tables
library(DT) #interactive table
library(ggrain)
library(chisq.posthoc.test) #chisquared post hoc test 
library(rcompanion)

#extra packages for plotting 
library(nord)
library(scico)
library(wesanderson)
library(taylor)
library(viridis)
library(ggtext)     # for extra text options
library(ggdist)     # for extra geoms
library(patchwork)  #  composing multiple plots

#regression visualisation
library(easystats)

#set working directory
#setwd("../ABCDTraumaPuberty/analy")
```

Load in cleaned dataframe created in LGCA script for a three class solution (best fitting class solution). Remember, this is the unrelated sample. 
```{r load data}

data <- readRDS("../data/cleanData3Class5.0.rds")

```

###Data prep

We need to:
1. Make dummy variables for the trauma_cat and trauma_bin variables for lavaan and chi-squared post hoc tests 
2. Make factor variables for site_d_l and ethnicity and a factor trauma variable for table purposes
3. Use scale function (default settings) to centre and scale numeric variables (mean centred and standardised (divided by standard deviation))

```{r data prep}

#need to recode trauma variable as a dummy variable with three levels, 1=no trauma, 2= one traumatic event, 3 = 2 or more traumatic event 
data <- data %>% 
  mutate(trauma_dummy = as.numeric(trauma_cat))
head(data$trauma_dummy)

str(data)
#make necessary variables factors 
data$site_id_l <- as.factor(data$site_id_l)
data$race_ethnicity <- as.factor(data$race_ethnicity)
data$trauma_factor <- as.factor(data$trauma_cat)

#recode ethnicity
#1 = White; 2 = Black; 3 = Hispanic; 4 = Asian; 5 = Other
data$race_ethnicity <- recode_factor(data$race_ethnicity, 
                                             "1" = "White",
                                             "2" = "Black",
                                             "3" = "Hispanic",
                                             "4" = "Asian",
                                             "5" = "Other")

#make recode trauma factor variable for table 
data$trauma_factor <- recode_factor(data$trauma_factor,
                                  "0" = "0 traumatic events",
                                  "1" = "1 traumatic event",
                                  "2" = "≥2 traumatic events")

```

##Descriptives table 
Make descriptives table with long format data with participants with complete trauma data
```{r descriptives with complete trauma data}

dataLongTbl <- data %>% 
  filter(!is.na(trauma_cat))

#remove people with missing trauma data
descTableMS <- dataLongTbl %>% 
  select(c(age_years, BPM_T4, trauma_factor, time, pds_avg, race_ethnicity, education_cat, demo_comb_income_v2)) %>% 
  tbl_summary(
    by = time,
    statistic = list(
      all_continuous() ~ c("{mean} ({sd})"),
      all_categorical() ~ c("{n} / {N} ({p}%)")
    ),  # <-- Closing parenthesis was missing here
    digits = list(
      all_continuous() ~ c(2, 2, 0, 0),
      all_categorical() ~ c(0, 0, 1)
    ),  # <-- Corrected the typo
    label = c(
      age_years ~ "Age in years",
      BPM_T4 ~ "Internalizing Symptoms (BPM)", 
      trauma_factor ~ "Trauma exposure",
      time ~ "time point",
      pds_avg ~ "PDS average score",
      race_ethnicity ~ "Ethnicity",
      education_cat ~ "Parental Education",
      demo_comb_income_v2 ~ "Household income"
    ),
    missing_text = "Missing",
    type = list(all_continuous() ~ "continuous"),
    missing = "ifany"
  ) %>% 
  bold_labels() %>% 
  italicize_levels()
descTableMS

```

##Tidy data
### Reduce dataframe

We want to tidy dataframe so that we have the variables we need for each participant. Note that we should have a dataframe with one row for each participant. Because of the input required for the LCGA, the data is in long format. 

To do, keep row for participant when time = 4, this should give us 3579 obs with the correct age_years variable. We can remove the other variables we don't need (e.g., pds_avg and pds_tot) as they are longitudinal and we have repeated measures for each participant. We don't need these data at present. 

```{r tidy df}
#reduce to variables of we need for main analysis 

df <- data %>% 
  select(c(id, src_subject_id, BPM_T4, trauma_cat, trauma_factor,trauma_dummy, trauma_T1, time, pds_avg, pds_tot, age_years, age_centred, site_id_l, bmi_score, race_ethnicity, education_cat, demo_comb_income_v2, slope, intercept, class, class_label))

#keep rows where time = 4, which should give us the correct dataframe 
df <- df %>%
  filter(time == 4) #3668

#check for duplicate IDs
uniqueIDs <- unique(df$src_subject_id) 
length(uniqueIDs) # N = 3668 - looks good! 

#rename age_years variable so it is clear that it is the age at T4
df <- df %>% 
  rename(age_T4 = age_years)

#our later analysis with lavaan requires complete data for the exogenous variable (trauma) so we will remove people with missing trauma data 
df <- df %>% 
  filter(!is.na(trauma_cat))

#scale numeric variables and save as new df to use later if needed
#Note: age has already been centred on T1 mean age so we do not need to scale here. 
dfScaled <- df
dfScaled$BPM_T4 <- scale(dfScaled$BPM_T4)
dfScaled$trauma_T1 <- scale(dfScaled$trauma_T1)
dfScaled$slope <- scale(dfScaled$slope)
dfScaled$intercept <- scale(dfScaled$intercept)

```

####Trauma/Puberty

```{r visualise trauma groups by puberty classes}
#check for group differences between trauma groups and puberty classification 

#change order of puberty classes for plotting purposes
df$class_label <- as.factor(df$class_label)
df$class_label <- factor(df$class_label, levels = c("Early starters", "Typical developers", "Slow developers"))

# Now, create the plot with the modified order
traumaPubertyBarplot <- ggplot(df, aes(x = class_label, fill = trauma_factor)) +
  geom_bar(position = "fill") +
  labs(
    title = "Trauma exposure and puberty class membership",
    x = "Puberty class",
    y = "Proportion of class"
  ) +
  scale_fill_manual(values = wesanderson::wes_palette("AsteroidCity1"),
                      guide = guide_legend(override.aes = list(size = 6))) +
  theme_classic() +
  theme(legend.title = element_blank()) +
  theme(legend.text = element_text(size = 16)) +
  theme(axis.text.x = element_text(size = 14, hjust = 0.5)) +
  theme(axis.text.y = element_text(size = 14, hjust = 1)) +
  theme(axis.title = element_text(size = 16, hjust = 0.5)) +
  theme(plot.title = element_text(size = 24, face = "bold", hjust = 0.5)) +
#try to add count 
  geom_text(
    aes(label = after_stat(count)),
    stat = "count",
    position = position_fill(vjust = 0.5),
    size = 5,
    color = "white"
  )

traumaPubertyBarplot
  
  
  
# Create the plot with custom legend labels
traumaPubertyBarplot <- ggplot(df, aes(x = as.factor(class_label), fill = trauma_factor)) +
  geom_bar(position = "fill") +
  labs(
    title = "Trauma exposure and puberty class membership",
    x = "Puberty class",
    y = "Proportion of class"
  ) +
  scale_fill_manual(values = c("#00b9e3", "#e88526" ,"#93aa00"),
                      guide = guide_legend(override.aes = list(size = 6))) +
  theme_classic() +
  theme(legend.title = element_blank()) +
  theme(legend.text = element_text(size = 16)) +
  theme(axis.text.x = element_text(size = 14, hjust = 0.5)) +
  theme(axis.text.y = element_text(size = 14, hjust = 1)) +
  theme(axis.title = element_text(size = 16, hjust = 0.5)) +
  theme(plot.title = element_text(size = 24, face = "bold", hjust = 0.5)) +
#try to add count 
  geom_text(
    aes(label = after_stat(count)),
    stat = "count",
    position = position_fill(vjust = 0.5),
    size = 5,
    color = "white"
  )
traumaPubertyBarplot
#save as .png
ggsave("../figs/traumaPubertyBarplot.png", plot = traumaPubertyBarplot, width = 10, height = 8, units = "in", dpi = 300)

```
#Hypothesis 1
### ANOVA: Puberty and trauma exposure 

```{r ANOVA puberty trauma}

levene_test(trauma_dummy ~ class_label, data = df) #homogeneity of variance assumption violated 
qqnorm(df$trauma_dummy) 
shapiro.test(df$trauma_dummy) #normality assumption violated also


kruskal.test(trauma_dummy ~ class_label, data = df) #p<0.001, suggests there is a significant difference between groups

#posthoc Dunn Test for pairwise comparison 
library(dunn.test)
dunn.test(df$trauma_dummy, g=df$class_label, method="bonferroni", list = TRUE)

#frequency table for trauma 
trauma_frq <- df%>% 
  group_by(class_label) %>% 
  get_summary_stats(trauma_dummy, show = c("mean", "sd", "ci"))
trauma_frq

```
### ANOVA: Puberty and depression 

Test whether puberty class membership is associated with depression 
Normality assumptions were not met for ANOVA test so non-parametric equivalents were carried out. 

```{r ANOVA puberty depression}

levene_test(BPM_T4 ~ class_label, data = df) #homogeneity of variance assumption violated 
qqnorm(df$BPM_T4) 
shapiro.test(df$BPM_T4) #normality assumption violated also

kruskal.test(BPM_T4 ~ class_label, data = df) #p<0.001, suggests there is a significant difference between groups

#posthoc Dunn Test for pairwise comparison 
library(dunn.test)
dunn.test(df$BPM_T4, g=df$class_label, method="bonferroni")

#frequency table for depression 
depression_frq <- df%>% 
  group_by(class_label) %>% 
  get_summary_stats(BPM_T4, show = c("mean", "sd", "ci"))
depression_frq


```

###Test direct effect

```{r trauma dep lm}

#without transforming - really non normal residuals. 
modCatNoTrans <- lmer(BPM_T4 ~ trauma_cat + age_T4 + (1|site_id_l), data = dfScaled)
summary(modCatNoTrans)

#look at residuals of non transformed model
resid <- residuals(modCatNoTrans)
hist(resid) #look right skewed
shapiro_test(resid) #indicates that the data is not normal. 

#make nicer histogram for paper
library(viridis)
untransBPMPlot <- ggplot(data = data.frame(resid = resid), aes(x = resid)) +
  geom_histogram(binwidth = .8, fill = viridis(1)) +
  labs(title = "Untransformed BPM total score ", x = "Residuals", y = "Frequency") +
  theme_minimal() +
  scale_fill_viridis(option = "magma") +
  scale_color_viridis(option = "magma") +
   xlim(-4, 10)
ggsave("../figs/UntransformedBPMTraumaHist.png", plot = untransBPMPlot, width = 6, height = 6, units = "in", dpi = 300)

#transform data using BestNormalize package. 
yeojohnson_obj <- yeojohnson(dfScaled$BPM_T4)
dfScaled$BPM_T4Trans <- predict(yeojohnson_obj)
#rerun  model 
modCatTrans <- lmer(BPM_T4Trans ~ trauma_cat + age_T4 + (1|site_id_l), data = dfScaled)
summary(modCatTrans)

#Check model residuals 
resid <- residuals(modCatTrans)
hist(resid) #residuals look much better

#make nice histogram 
transBPMPlot <- ggplot(data = data.frame(resid = resid), aes(x = resid)) +
  geom_histogram(binwidth = .8, fill = viridis(1)) +
  labs(title = "Transformed BPM total score ", x = "Residuals", y = "Frequency") +
  theme_minimal() +
  scale_fill_viridis(option = "magma") +
  scale_color_viridis(option = "magma") +
   xlim(-4, 10)
ggsave("../figs/TransformedBPMTraumaHist.png", plot = transBPMPlot, width = 6, height = 6, units = "in", dpi = 300)

shapiro_test(resid) #still says not normal but large sample size should comply with central limit theory, so it should be okay! 



```


##H2: Mediation 

We will use the lavaan package to test for mediation.

We can use bootstraping to account for the non-normal distribution of the data. For our main analysis, we will use a continuous trauma variable, and a continuous depression variable. 

Parallel mediation = slope and intercept

The parallel mediation model will tell us whether, regardless of pubertal status at baseline (the intercept), the rate at which you progress through puberty (the slope), mediates the association between trauma exposure and depression.

We will use the lavaan package to test for mediation. 
Paths in model:

a = trauma to mediator
b = depression to mediator
c = trauma to depression 

Graph_sem tutorial: https://cjvanlissa.github.io/tidySEM/articles/sem_graph.html

Bootstrapping = 5000 interations. 
```{r lavaan model}
#load in lavaan model generated from code below 
modelMulti <- readRDS("/Users/niamhmacsweeney/Library/CloudStorage/OneDrive-UniversityofEdinburgh/Edinburgh/ABCD_collabs/ABCD_trauma_puberty_dep/data/lavaanMultipleMediationOutput.rds")

#multiple mediation model 
#rename variables so they are clearer when plotted
dfScaled <- dfScaled %>% 
  rename(interc = intercept, #rename intercept so that lavaan works
         trauma = trauma_dummy,
         intSx = BPM_T4Trans,
         age = age_centred,
         BMI = bmi_score,
         ethnicity = race_ethnicity,
         income = demo_comb_income_v2,
         education = education_cat)
         

#check covariance between slope and intercept
cov(df$slope, df$intercept, method = "pearson") #not correlated

modelMulti <- '
          # multiple mediation model with slope and intercept
             slope ~ a1*trauma + age
             interc ~ a2*trauma + age
             intSx ~ b1*slope + b2*interc + c*trauma + age
          #  covariance between intercept and slope 
             slope ~~ interc
           # indirect effect (a*b)
             indirect1 := a1*b1
             indirect2 := a2*b2
           # total effect
             total := c + (a1*b1) + (a2*b2) 
           # direct effect
             direct := c 
          '
set.seed(2507)
fitMulti <- sem(modelMulti, missing = "ML", meanstructure = TRUE, se = "bootstrap", bootstrap = 5000, data = dfScaled, fixed.x = FALSE)
summary(fitMulti,  fit.measures = TRUE, standardized=TRUE)
# get bias corrected standard errors from the bootstrap
parameterestimates(fitMulti , boot.ci.type = "bca.simple")

library(lavaanPlot)
test <- lavaanPlot(model = fitMulti, node_options = list(shape = "box", fontname = "Helvetica"), edge_options = list(color = "grey"),covs = TRUE, coefs = TRUE, sig = .05, stars = "regress")
test
modelMultiGraph <- graph_sem(model = fitMulti)
modelMultiGraph
#save lavaan output with bootstrapping so we don't have to re-run
# saveRDS(modelMulti, "/Users/niamhmacsweeney/Library/CloudStorage/OneDrive-UniversityofEdinburgh/Edinburgh/ABCD_collabs/ABCD_trauma_puberty_dep/data/lavaanMultipleMediationOutput.rds")

```

##Intercept only model 
```{r intercept mediation}

modelInter <- '
          # mediation model with intercept
             interc ~ a1*trauma + age
             intSx ~ b1*interc + c*trauma + age
           # indirect effect (a*b)
             indirect1 := a1*b1
           # total effect
             total := c + (a1*b1)
           # direct effect
             direct := c 
          '
set.seed(2507)
fitInter <- sem(modelInter, missing = "ML", meanstructure = TRUE, se = "bootstrap", bootstrap = 5000, data = dfScaled)
summary(fitInter, fit.measures = TRUE, standardized=TRUE)
# get bias corrected standard errors from the bootstrap
parameterestimates(fitInter, boot.ci.type = "bca.simple")
modelInterGraph <- graph_sem(model = fitInter)
modelInterGraph
```

##Slope only model 
```{r slope mediation}

modelSlope <- '
          # mediation model with slope
             slope ~ a1*trauma + age
             intSx ~ b1*slope + c*trauma + age
           # indirect effect (a*b)
             indirect1 := a1*b1
           # total effect
             total := c + (a1*b1)
           # direct effect
             direct := c 
          '
set.seed(2507)
fitSlope <- sem(modelSlope, missing = "ML", meanstructure = TRUE, se = "bootstrap", bootstrap = 5000, data = dfScaled)
summary(fitSlope, fit.measures = TRUE, standardized=TRUE)
# get bias corrected standard errors from the bootstrap
parameterestimates(fitSlope, boot.ci.type = "bca.simple")
modelSlopeGraph <- graph_sem(model = fitSlope, reor)
modelSlopeGraph
```





##Sensitivity analysis 

###Covariates
Explore the impact of BMI and household income on mediation model. 

```{r prep data for sen analysis}

#we will household income as a continuous variable as it is ordered

dfScaled$income_num <- as.numeric(dfScaled$income)
dfScaled$education_num <- as.numeric(dfScaled$education)
#we will also need to scale BMI 
dfScaled$BMI_std <- scale(dfScaled$BMI)

```

```{r mediation with covs}

modelMultiCov <- '
  # multiple mediation model with slope and intercept
  slope ~ a1*trauma + age + income_num + BMI_std
  interc ~ a2*trauma + age + income_num + BMI_std
  intSx ~ b1*slope + b2*interc + c*trauma + age + income_num + BMI_std 
  
  # covariance of mediators
  slope ~~ interc
  
  # indirect effect (a*b)
  indirect1 := a1*b1
  indirect2 := a2*b2
  
  # total effect
  total := c + (a1*b1) + (a2*b2)
  
  # direct effect
  direct := c
'

set.seed(2507)
fitCov <- sem(modelMultiCov, missing = "ML", meanstructure = TRUE, se = "bootstrap", bootstrap = 5000, data = dfScaled, fixed.x = FALSE) #fixed.x = F means that the covariates are considered random, and the means, variances and covariances are free parameters. This allows us to use ML to handle missing data still. 
summary(fitCov,  fit.measures = TRUE, standardized=TRUE)
# get bias corrected standard errors from the bootstrap
parameterestimates(fitCov, boot.ci.type = "bca.simple")
fitCovGraph <- graph_sem(model = fitCov)
fitCovGraph

```

###Un-coded trauma variable

Re-run mediation analysis with un-coded trauma variable as a continuous predictor 



```{r un-coded trauma variable}

#multiple mediation model 
#rename variables so they are clearer when plotted
dfScaled <- dfScaled %>% 
  rename(interc = intercept, #rename intercept so that lavaan works
         traumaR = trauma_dummy,
         intSx = BPM_T4Trans,
         age = age_centred,
         BMI = bmi_score,
         ethnicity = race_ethnicity,
         income = demo_comb_income_v2,
         education = education_cat)
        

modelMultiTraumaR <- '
          # multiple mediation model with slope and intercept
             slope ~ a1*traumaR + age
             interc ~ a2*traumaR + age
             intSx ~ b1*slope + b2*interc + c*traumaR + age
          #  covariance between intercept and slope 
             slope ~~ interc
           # indirect effect (a*b)
             indirect1 := a1*b1
             indirect2 := a2*b2
           # total effect
             total := c + (a1*b1) + (a2*b2) 
           # direct effect
             direct := c 
          '
set.seed(2507)
fitTraumaR <- sem(modelMultiTraumaR, missing = "ML", meanstructure = TRUE, se = "bootstrap", bootstrap = 5000, data = dfScaled, fixed.x = FALSE)
summary(fitTraumaR ,  fit.measures = TRUE, standardized=TRUE)
# get bias corrected standard errors from the bootstrap
parameterestimates(fitTraumaR  , boot.ci.type = "bca.simple")

```












Try and re-run model without age --- could include if asked for at later stage. Results don't change. 

Results are the same
```{r multi med without age}

modelMulti2 <- '
          # multiple mediation model with slope and intercept
             slope ~ a1*trauma
             interc ~ a2*trauma
             intSx ~ b1*slope + b2*interc + c*trauma
          #  covariance of mediators
             slope ~~ interc
           # indirect effect (a*b)
             indirect1 := a1*b1
             indirect2 := a2*b2
           # total effect
             total := c + (a1*b1) + (a2*b2)
           # direct effect
             direct := c
          '
set.seed(2507)
fitMulti2 <- sem(modelMulti2, missing = "ML", meanstructure = TRUE, se = "bootstrap", bootstrap = 5000, data = dfScaled)
summary(fitMulti2,  fit.measures = TRUE, standardized=TRUE)
parameterestimates(fitMulti2 , boot.ci.type = "bca.simple")
modelMulti2Graph <- graph_sem(model = fitMulti2)
modelMulti2Graph


```


##Exploratory analysis
###Pubertal timing

We will be using the "data" object that was loaded at the start of this script, and includes repeated measures for each participant ["cleanData3Class5.0.rds"]

```{r view data}

view(data)

```

####Calculate pubertal timing score 

To generate a measure of pubertal timing we will follow the method used by N. Vijayakumar et al. (2023) [https://www.cambridge.org/core/product/identifier/S0033291723001472/type/journal_article], where pubertal timing was calculated using the following linear mixed model:

linear mixed model: PDS_avg ∼ age + (1|subject_id) + (1| site_id).

Update = don't include subject random effect - just the residual. 
We will then sum the subject specific random effect and the model residuals as the measure of pubertal timing, which reflect ‘stable’ intercept differences per individual (across time points) in addition to residual differences at each time point. This allowed us to estimate a unique value for pubertal timing for each time point. 

[Text above from Vijayakumar et al., 2023 paper]

Note: We will not include age^2 as a covariate in the model so that we are consistent with the models in our earlier analysis, which differs from the Vijayakumar et al. paper. 

We will use the ranef function from lmer to extract the subject specific random effect. See documentation here: https://rdrr.io/cran/lme4/man/ranef.html

```{r calculate pubertal timing score}

data$id <- as.factor(data$id)
ptMod <- lmer(pds_avg ~ age_centred + (1|id) + (1|site_id_l), data = data)
summary(ptMod)
str(ptMod) #gives the structure of the model 

#1 Extract subject specific random effect (this will be the same for each unique participant across timepoints). This is their intercept. 

#need to filter so we have random effects for id only 
ranefDF <- as.data.frame(ranef(ptMod)) %>% 
  filter(grpvar == "id")

#save subject specific random effect as new variable in main dataframe (data)
ranefDF <- ranefDF %>% 
  rename(id = grp) #rename column so it is the same across both dataframes
data <- merge(data, ranefDF[, c("id", "condval")], by = "id") #merge by id based on indexing 

#rename condval to sub_ranef
data <- data %>% 
  rename(sub_ranef = condval)

#2 Extract residuals (this will differ for each participant across timepoints)
data$residuals <- resid(ptMod) #given that order of rows is the same, we can just extract resids and add to dataframe

#3 Pubertal timing = residuals 

data <- data %>% 
  mutate(pt = residuals) #looks good

```

####Make age groups
We want to group the data into age bins so that we can re-run our mediation analysis within different age groups. This will allow us to test whether there are ages at which pubertal timing might differ in how it mediates the association between early life trauma and later depression. 

We will first group participants based on six month age bins.  

1. 9 to < 9.5 years
2. 9.5 to < 10 years
3. 10 to < 10.5 years 
4. 10.5 to < 11 years
5. 11 to < 11.5 years 
6. 11.5 to < 12 years
7. 12 to < 12.5 years 
8. 12.5 to < 13 years
9. 13 to < 13.5 years 
10. 13.5 to < 14 years 

We will remove participants that fall below the lower limit (N = 57) and upper limit (N = 84) of the first age and last age groups so that the age groupings and sizes are consistent. This will remove 141 IDs in total. 

```{r define age groups}

data <- data %>% 
  filter(!(data$age_years < 9 | data$age_years >14)) #should remove 144 IDs (N = 15316 obs)

data <- data %>% 
mutate(age_group = case_when(
      age_years >= 9 & age_years < 9.5 ~ "1",
      age_years >= 9.5 & age_years < 10 ~ "2",
      age_years >= 10 & age_years < 10.5 ~ "3",
      age_years >= 10.5 & age_years < 11 ~ "4",
      age_years >= 11 & age_years < 11.5 ~ "5",
      age_years >= 11.5 & age_years < 12 ~ "6",
      age_years >= 12 & age_years < 12.5 ~ "7",
      age_years >= 12.5 & age_years < 13 ~ "8",
      age_years >= 13 & age_years < 13.5 ~ "9",
      age_years >= 13.5 & age_years <= 14 ~ "10"
)) #this looks like it worked!


#convert character to numeric 
data$age_group<- as.numeric(data$age_group)

table(data$age_group) #groups look fairly even

```
####Check for duplicate IDs in age groups 

It seems that there are duplicate IDs in groups 6,7,8,9,10. 

```{r find duplicate IDs}
# Group the data by age_group
grouped_data <- data %>% group_by(age_group)

# Identify duplicate IDs within each age group
duplicates <- grouped_data %>%
  filter(n_distinct(src_subject_id) != n()) %>%
  ungroup()

# Create a data frame with all observations of the duplicate IDs within their respective groups
all_duplicate_observations <- data %>%
  filter(age_group %in% duplicates$age_group, src_subject_id %in% duplicates$src_subject_id)

# Filter to only return IDs with the same src_subject_id and age_group
final_duplicate_observations <- all_duplicate_observations %>%
  group_by(src_subject_id, age_group) %>%
  filter(n() > 1) %>%
  ungroup()

# View the resulting data frame
print(final_duplicate_observations)

# #reduce to main variables for inspection purposes
# final_duplicate_observations <- final_duplicate_observations %>% 
#   select(c(src_subject_id, age_years, age_group, time))
```

Given that there are duplicate ids for some participants within the 6-month age groups, we will randomly select an entry per participant so that we only have one observation for each participant in each age group. The intervals between the "yearly" study visits seem very inconsistent for some participants. For example, the follow up time is only two months for some participants! Maybe this is something to consider for future work. How to deal with the variability in follow up timepoints? 

```{r tidy duplicate IDs}

# Randomly select one observation per participant within their age_group
set.seed(2507)
selected_observations <- final_duplicate_observations %>%
  group_by(age_group, src_subject_id) %>%
  sample_n(1) %>%
  ungroup()

#remove ids in selected_observations (n=35)
# Filter here by id and age_years (which is different across duplicate ids) as age_group is not.
#should have 15291 obs in df
data_clean <- data %>%
  anti_join(selected_observations, by = c("age_years", "src_subject_id"))

```


We will need to scale the new numeric variables we have added to the dataframe, and then replace the old dfScaled df
####Prep variables for lavaan
```{r scale new variables}


data <- data %>% 
  mutate(trauma_dummy = as.numeric(trauma_cat))
head(data$trauma_dummy)



#scale non dummy numeric variables and save as new df
dfScaled <- data
dfScaled$BPM_T4 <- scale(dfScaled$BPM_T4)
dfScaled$slope <- scale(dfScaled$slope)
dfScaled$intercept <- scale(dfScaled$intercept)
dfScaled$sub_ranef <- scale(dfScaled$sub_ranef)
dfScaled$trauma_T1 <- scale(dfScaled$trauma_T1)
dfScaled$pt <- scale(dfScaled$pt)
view(dfScaled)

```

####Age group mediation

The mediation model set up is the same as our main mediation analysis but the mediator this time is pubertal timing

M = pubertal timing score
X = Trauma 
Y = Depression

```{r define lavaan model}

modelPT <- '
          # mediation model
             pt ~ a1*trauma_dummy
             BPM_T4 ~ b1*pt + c*trauma_dummy
           # indirect effect (a*b)
             indirect1 := a1*b1
           # total effect
             total := c + (a1*b1)
           # direct effect
             direct := c 
          '

```

Get updated table counts per age group after removing duplicates
```{r age group table}

age_group_counts <- table(dfScaled$age_group)
view(age_group_counts)

```

Group 1 (9 to 9.5 years old) N = 1198
```{r group 1}

#filter data
grp1 <- dfScaled %>% 
  filter(age_group == 1)


set.seed(2507)
fitGrp1 <- sem(modelPT, missing = "ML", meanstructure = TRUE, se = "bootstrap", bootstrap = 1000, data = grp1, fixed.x = FALSE)
summary(fitGrp1,  fit.measures = TRUE, standardized=TRUE)
# get bias corrected standard errors from the bootstrap
grp1Est <- parameterestimates(fitGrp1, boot.ci.type = "bca.simple")
#extract vals of interest
grp1Vals <- grp1Est[10:12,]
grp1Vals$Group <- c(1,1,1)

fitGrp1Graph <- graph_sem(model = fitGrp1) #plot
fitGrp1Graph
```

Group 2 (9.5 to 10 years old) N = 1066
```{r group 2}

#filter data
grp2 <- dfScaled %>% 
  filter(age_group == 2)

fitGrp2 <- sem(modelPT, missing = "ML", meanstructure = TRUE, se = "bootstrap", bootstrap = 1000, data = grp2, fixed.x = FALSE)
summary(fitGrp2,  fit.measures = TRUE, standardized=TRUE)
# get bias corrected standard errors from the bootstrap
grp2Est <- parameterestimates(fitGrp2, boot.ci.type = "bca.simple")
#extract vals of interest
grp2Vals <- grp2Est[10:12,]
grp2Vals$Group <- c(2,2,2)

fitGrp2Graph <- graph_sem(model = fitGrp2) #plot
fitGrp2Graph

```
Group 3 (10 to 10.5 years old) N = 1956
```{r group 3}

grp3 <- dfScaled %>% 
  filter(age_group == 3)

fitGrp3 <- sem(modelPT, missing = "ML", meanstructure = TRUE, se = "bootstrap", bootstrap = 1000, data = grp3, fixed.x = FALSE)
summary(fitGrp3,  fit.measures = TRUE, standardized=TRUE)
# get bias corrected standard errors from the bootstrap
grp3Est <- parameterestimates(fitGrp3, boot.ci.type = "bca.simple")
#extract vals of interest
grp3Vals <- grp3Est[10:12,]
grp3Vals$Group <- c(3,3,3)

fitGrp3Graph <- graph_sem(model = fitGrp3) #plot
fitGrp3Graph


```
Group 4 (10.5 to 11 years old) N = 1919

```{r group 4}

grp4 <- dfScaled %>% 
  filter(age_group == 4)

fitGrp4 <- sem(modelPT, missing = "ML", meanstructure = TRUE, se = "bootstrap", bootstrap = 1000, data = grp4, fixed.x = FALSE)
summary(fitGrp4,  fit.measures = TRUE, standardized=TRUE)
# get bias corrected standard errors from the bootstrap
grp4Est <- parameterestimates(fitGrp4, boot.ci.type = "bca.simple")
#extract vals of interest
grp4Vals <- grp4Est[10:12,]
grp4Vals$Group <- c(4,4,4)

fitGrp4Graph <- graph_sem(model = fitGrp4) #plot
fitGrp4Graph


```
Group 5 (11 to 11.5 year olds) N = 1771
```{r group 5}

grp5 <- dfScaled %>% 
  filter(age_group == 5)

fitGrp5 <- sem(modelPT, missing = "ML", meanstructure = TRUE, se = "bootstrap", bootstrap = 1000, data = grp5, fixed.x = FALSE)
summary(fitGrp5,  fit.measures = TRUE, standardized=TRUE)
# get bias corrected standard errors from the bootstrap
grp5Est <- parameterestimates(fitGrp5, boot.ci.type = "bca.simple")
#extract vals of interest
grp5Vals <- grp5Est[10:12,]
grp5Vals$Group <- c(5,5,5)

fitGrp5Graph <- graph_sem(model = fitGrp5) #plot
fitGrp5Graph
```

Group 6 (11.5 to 12 year olds) N = 1918
```{r group 6}

grp6 <- dfScaled %>% 
  filter(age_group == 6)

fitGrp6 <- sem(modelPT, missing = "ML", meanstructure = TRUE, se = "bootstrap", bootstrap = 1000, data = grp6, fixed.x = FALSE)
summary(fitGrp6,  fit.measures = TRUE, standardized=TRUE)
# get bias corrected standard errors from the bootstrap
grp6Est <- parameterestimates(fitGrp6, boot.ci.type = "bca.simple")
#extract vals of interest
grp6Vals <- grp6Est[10:12,]
grp6Vals$Group <- c(6,6,6)

fitGrp6Graph <- graph_sem(model = fitGrp6) #plot
fitGrp6Graph
```

Group 7 (12 to 12.5 year olds) N = 1948
```{r group 7}

grp7 <- dfScaled %>% 
  filter(age_group == 7)

fitGrp7 <- sem(modelPT, missing = "ML", meanstructure = TRUE, se = "bootstrap", bootstrap = 1000, data = grp7, fixed.x = FALSE)
summary(fitGrp7,  fit.measures = TRUE, standardized=TRUE)
# get bias corrected standard errors from the bootstrap
grp7Est <- parameterestimates(fitGrp7, boot.ci.type = "bca.simple")
#extract vals of interest
grp7Vals <- grp7Est[10:12,]
grp7Vals$Group <- c(7,7,7)

fitGrp7Graph <- graph_sem(model = fitGrp7) #plot
fitGrp7Graph
```

Group 8 (12.5 to 13 year olds) N = 1623
```{r group 8}

grp8 <- dfScaled %>% 
  filter(age_group == 8)

fitGrp8 <- sem(modelPT, missing = "ML", meanstructure = TRUE, se = "bootstrap", bootstrap = 1000, data = grp8, fixed.x = FALSE)
summary(fitGrp8,  fit.measures = TRUE, standardized=TRUE)
# get bias corrected standard errors from the bootstrap
grp8Est <- parameterestimates(fitGrp8, boot.ci.type = "bca.simple")
#extract vals of interest
grp8Vals <- grp8Est[10:12,]
grp8Vals$Group <- c(8,8,8)

fitGrp8Graph <- graph_sem(model = fitGrp8) #plot
fitGrp8Graph


```

Group 9 (13 to 13.5 years) N = 1119
```{r group 9}

grp9 <- dfScaled %>% 
  filter(age_group == 9)

fitGrp9 <- sem(modelPT, missing = "ML", meanstructure = TRUE, se = "bootstrap", bootstrap = 1000, data = grp9, fixed.x = FALSE)
summary(fitGrp9,  fit.measures = TRUE, standardized=TRUE)
# get bias corrected standard errors from the bootstrap
grp9Est <- parameterestimates(fitGrp9, boot.ci.type = "bca.simple")
#extract vals of interest
grp9Vals <- grp9Est[10:12,]
grp9Vals$Group <- c(9,9,9)

fitGrp9Graph <- graph_sem(model = fitGrp9) #plot
fitGrp9Graph


```
Group 10 (13.5 to 14 years) N = 773

```{r group 10}

grp10 <- dfScaled %>% 
  filter(age_group == 10)

fitGrp10 <- sem(modelPT, missing = "ML", meanstructure = TRUE, se = "bootstrap", bootstrap = 1000, data = grp10, fixed.x = FALSE)
summary(fitGrp10,  fit.measures = TRUE, standardized=TRUE)
# get bias corrected standard errors from the bootstrap
grp10Est <- parameterestimates(fitGrp10, boot.ci.type = "bca.simple")
#extract vals of interest
grp10Vals <- grp10Est[10:12,]
grp10Vals$Group <- c(10,10,10)
fitGrp10Graph <- graph_sem(model = fitGrp10) #plot
fitGrp10Graph


```

####Results table {.tabset}

```{r age group results}
grp1Vals #not sig
grp2Vals #not sig
grp3Vals #not sig after multiple comparison correction (MCC)
grp4Vals #sig total and direct after MCC
grp5Vals #not sig after MCC
grp6Vals #not sig
grp7Vals #sig total and direct after MCC
grp8Vals #not sig
grp9Vals #not sig after MC
grp10Vals #not sig

#bind dfs together to make a table 
resultsExpl <- rbind(grp1Vals, grp2Vals, grp3Vals, grp4Vals, grp5Vals, grp6Vals, grp7Vals, grp8Vals, grp9Vals, grp10Vals)

```

####Timing at each timepoint 

Mabe look at this later 

First run mediation model with pubertal timing as the mediator at each follow up time point (T1 to T4)

There is a significant mediation effect of pubertal timing on the relationship between trauma and depression
```{r pubertal timing as mediator}

modelPT <- '
          # mediation model
             pt ~ a*trauma_dummy3 + rane
             BPM_T4 ~ b*pt + c*trauma_dummy3
           # indirect effect (a*b)
             ab := a*b
           # total effect
             total := c + (a*b) '

fitPT <- sem(modelPT, missing = "ML", meanstructure = TRUE, se = "bootstrap", bootstrap = 5000, data = dfScaled)
summary(fitPT,  fit.measures = TRUE, standardized=TRUE)
# get bias corrected standard errors from the bootstrap
parameterestimates(fitPT , boot.ci.type = "bca.simple")

modelPTGraph <- graph_sem(model = fitPT) #plot


```


